Speeding Up Sequence Alignment Algorithms via Parallel Programming: Dividing Work into Chunks

### 2016-06-06 paper.tex
Final version for draft.
Changes from papeeria

The dynamic programming algorithms essentially divides a large problem (e.g. the full sequence) into a series of smaller problems and uses the optimal solutions to the smaller problems to reconstruct an optimal solution to the larger problem.

In 1970, Saul B. Needleman y Christian D. Wunsch published \cite{needleman1970align}, a method for global sequence alignment with dynamic programming (one table), and in 1981, Temple F. Smith y Michael S. Waterman published \cite{smith1981align}, a method for local alignment of sequences consisting of a Needleman-Wunsch variant method. However, these methods did not have mechanisms for compaction of the sequences.

@@

As future work we propose implementing other approaches and running the same experiment as we did, to be able to compare statistically our results. One of this other approaches, that has not escapes our mind, is trying different ways of synchronization in our chunk-based solution. Currently, the thread management is mainly done via semaphores. Another option would be to use a \textit{yield}-like solution. Meaning that a thread relinquish its CPU time directly, without having the semaphores as intermediary. For that there are multiple c primitives to aid the programmer in such circumstances such as: \textit{pthread\_yield} or \textit{sched\_yield} that causes the calling thread to relinquish the CPU to another thread; also we could use something such as \textit{sleep} which makes sure that the calling thread will not run on the CPU for a given amount of time.

>>>>

The dynamic programming algorithms essentially divides a large problem (e.g. the full sequence) into a series of smaller problems and uses the optimal solutions to the smaller problems to reconstruct an optimal solution to the larger problem. This is known as the ``Principle of Optimallity'' \cite{rbellman1957}.

A Bellman Equation describes the policy that the algorithm should use to make an optimal decision from the smaller sub-optimal decisions \cite{rbellman1957}.

In 1970, Saul B. Needleman y Christian D. Wunsch published \cite{needleman1970align}, a method for global sequence alignment with dynamic programming (one table), and in 1981, Temple F. Smith y Michael S. Waterman published \cite{smith1981align}, a method for local alignment of sequences consisting of a Needleman-Wunsch variant method. However, these methods did not have mechanisms for compaction of the sequences. The bellman equation used by the Needleman-Wunsch algorithm is:

\begin{center}
    $H_{i,j} = max(H_{i-1,j-1} + M, H_{i-1,j} + GP, H_{i,j-1} + GP $ \cite{needleman1970align}
\end{center}

Where M is the gain of a match in the sequence and GP is the penalty of a gap, which can vary depending on the scheme used \cite{needleman1970align}. Similarly the Bellman Equation for the Smith-Waterman algorithm is:

\begin{center}
    $H_{i,j} = max(0, H_{i-1,j-1} + M, H_{i-1,j} + GP, H_{i,j-1} + GP) $ \cite{smith1981align}
\end{center}

It is different in that the minimum value that is going to be used is a 0, no negative values are allowed, since when $H_{i-1,j-1} + M$ and $H_{i-1,j} + GP$ and $H_{i,j-1} + GP$ are negative, the 0 will be the max value.

@@

As future work we propose implementing other approaches and running the same experiment as we did, to be able to compare statistically our results. One of this other approaches, that has not escaped our mind, is trying different ways of synchronization in our chunk-based solution. Currently, the thread management is mainly done via semaphores. Another option would be to use a \textit{yield}-like solution. Meaning that a thread relinquish its CPU time directly, without having the semaphores as intermediary. For that there are multiple c primitives to aid the programmer in such circumstances such as: \textit{pthread\_yield} or \textit{sched\_yield} that causes the calling thread to relinquish the CPU to another thread; also we could use something such as \textit{sleep} which makes sure that the calling thread will not run on the CPU for a given amount of time.
